### 2012-03-03 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

\subsection{PathType}
$PathType$ contains additional information. The represented path is contained by a list of its steps, in particular instances of $StepExprNode$. If a step is a reference to a variable which type is $PathType$, we also want to include this information. Therefore, $PathType$ contain association between the steps and other $PathTypes$ and this association is defined for the $PathType$ variable steps.

@@

In summary, we represent $PathType$ as a structure with the following member variables. \todo{Radsej UML?}
\begin{itemize}
\item $stepNodes$ - A list of references to $StepExprNode$ instances.
\item $forBoundSubsteps$ - An association between references from $stepNodes$ and instances of $PathType$ type.
\item $isForBound$ - Boolean flag determining if the type was bound to a for variable in a for clause.
\item $specialFunctionCalls$ - List of special functions called with this instance as an argument.
\end{itemize}

@@

Without loss of generality, we will focus on locally defined functions. We will not determine types of functions defined in other modules since the principle is similar but it is needed to look up the definitions in syntax trees of other queries. Return types of built-in functions are fixed and, thus, there is no need to analyze them.

@@

\STATE $type = UnknownType$
\IF{$is(exprNode, LiteralNode)$}
     \STATE $type =$ XSD atomic type $exprNode.type$, exactly-once cardinality
\ELSIF{$is(exprNode, FunctionCallNode)$}
     \STATE $type = getFunctionReturnType(exprNode.fncName)$
\ELSIF{$is(exprNode, VarRefNode)$}
     \STATE $type = getVariableType(contextVarTypes, exprNode.varName)$
\ELSIF{$is(exprNode, PathExprNode)$}
     \STATE $type = createPathType(exprNode)$
\ELSIF{$is(exprNode, OperatorNode)$}
     \STATE $type = determineOperatorType(exprNode, contextVarTypes)$
\ELSIF{$is(exprNode, FLWORExprNode)$}
     \STATE $returnClauseNode = exprNode.getChild(ReturnClauseNode)$
     \STATE $type = createForUnboundType(getType(returnClauseNode))$
\ENDIF

@@

The algorithm iterates through the variable declaration nodes from the prolog and if the variable is explicitly assigned with its type, the type is noted. Otherwise, an attempt of the type deduction of the binding expression is made. The deduction of the type from the expression is presented in Algorithm \ref{ALG_func_determineExpressionType}. On input, it takes an expression node and types of local variables that are valid in the expression (also called context of variables or variable context). The presented algorithm is just shortened illustration since the complete version is too long to be presented in this text. Nevertheless, the principle is straightforward. Depending on the class of the expression node (show in Figure \ref{FIG_syntax_tree_node_types_part_1}), its type can be either determined directly or it depends on its subexpressions. A part of the function's semantic is separated in Algorithm \ref{ALG_func_determineOperatorType}, which is not presented completely because of the same reason.

@@

\IF{$is(node, PathExprNode)$}
	\STATE $type$ = $getType(node)$
	\IF{$type$ is $PathType$ AND $isTargetPath(type, varName)$}
		\STATE add $type$ to $paths$
	\ENDIF
\ENDIF
\FORALL{$child \in node.getChildren()$}
	\STATE $getTargetReturnPathsRecursive(child, varName, paths)$
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

The first group of XQuery constructs utilized by this step is a group of $distinct-values$ function and aggregation functions $min$, $max$, $sum$ as discussed in Section \ref{Negative_uniqueness} and shown in Algorithm \ref{ALG_rejection_of_uniqueness_aggregation_functions}. The algorithm searches for calls of the mentioned functions with a $PathType$ on their input. Elements and attributes selected by those $PathTypes$ are then considered to be not unique, and consequently, they cannot be keys.

@@

\IF{$is(node, FunctionCallNode)$}
    \IF{$node.fncName$ is one of $distinct-values$, $min$, $max$, $sum$}
        \STATE $argument = node.getChild(0)$
        \STATE $type = getType(argument)$
        \IF{$type$ is $PathType$ AND $usesOnlyChildAndDescendantAxes(type)$ AND $isWithoutPredicatesExceptLastStep(type)$}
            \STATE memorize $argument$ as the negative key statement, weight 1
        \ENDIF
    \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

>>>>

\subsection{PathType}
$PathType$ contains additional information. The represented path is contained by an instance of $PathExprNode$ containing list of its steps, in particular instances of $StepExprNode$. If a step is a reference to a variable which type is $PathType$, we also want to include this information. Therefore, $PathType$ contain association between the steps and other $PathTypes$ and this association is defined for the $PathType$ variable steps.

@@

In summary, we represent $PathType$ as a structure with the following member variables. \todo{Radsej UML?}
\begin{itemize}
\item $pathExprNode$ - An instance of $PathExprNode$.
\item $substeps$ - An association between references from $stepNodes$ and instances of $PathType$ type.
\item $isForBound$ - Boolean flag determining if the type was bound to a for variable in a for clause.
\item $specialFunctionCalls$ - List of special functions called with this instance as an argument.
\end{itemize}

@@

Without loss of generality, we will focus on locally defined functions. We will not determine types of functions defined in other modules since the principle is similar but it is needed to look up the definitions in syntax trees of other queries. Return types of built-in functions are fixed and, thus, there is no need to analyze them. Also, since a determination of a prefix for built-in functions is a technical issue, we assume that built-in functions are either prefixed by $fn$ or not prefixed.

@@

\STATE $type = UnknownType$
\IF{$is(exprNode, LiteralNode)$}
     \STATE $type =$ XSD atomic type $exprNode.type$, exactly-once cardinality
\ELSIF{$is(exprNode, FunctionCallNode)$}
     \STATE $type = getFunctionReturnType(exprNode.fncName)$
\ELSIF{$is(exprNode, VarRefNode)$}
     \STATE $type = getVariableType(contextVarTypes, exprNode.varName)$
     \IF{$type$ is $PathType$}
         \STATE $step$ = a path step (new instance of $StepExprNode$) representing the variable reference
		 \STATE $substeps$ = empty associative array         
         \STATE $substeps[step]$ = $type$
         \STATE $pathExprNode$ = new instance of $PathExprNode$ containing one step $step$
         \STATE $type$ = $PathType$ containing $pathExprNode$ and $substeps$
     \ENDIF
\ELSIF{$is(exprNode, PathExprNode)$}
     \STATE $type = createPathType(exprNode)$
\ELSIF{$is(exprNode, OperatorNode)$}
     \STATE $type = determineOperatorType(exprNode, contextVarTypes)$
\ELSIF{$is(exprNode, FLWORExprNode)$}
     \STATE $returnClauseNode = exprNode.getChild(ReturnClauseNode)$
     \STATE $type = createForUnboundType(getType(returnClauseNode))$
\ENDIF

@@

The algorithm iterates through the variable declaration nodes from the prolog and if the variable is explicitly assigned with its type, the type is noted. Otherwise, an attempt of the type deduction of the binding expression is made. The deduction of the type from the expression is presented in Algorithm \ref{ALG_func_determineExpressionType}. On input, it takes an expression node and types of local variables that are valid in the expression (also called context of variables or variable context). The presented algorithm is just shortened illustration since the complete version is too long to be presented in this text. Nevertheless, the principle is straightforward. Depending on the class of the expression node (show in Figure \ref{FIG_syntax_tree_node_types_part_1}), its type can be either determined directly or it depends on its subexpressions.

The meaning of the code at lines 6-14 is the following. If a variable is bound to an expression which type is $PathType$, we want to assign the variable reference expression with $PathType$ which represents path containing exactly one step, the variable reference. Some algorithm in later sections search for paths starting with a certain variable and we want them to include also that types of expressions.

A part of the function's semantic is separated in Algorithm \ref{ALG_func_determineOperatorType}, which is not presented completely because of the same reason.

@@

\IF{$is(node, PathExprNode)$}
	\STATE $type$ = $getType(node)$
	\IF{$type$ is $PathType$ AND $isTargetPath(type, varName)$}
		\STATE add $type$ to $paths$
	\ENDIF
\ELSIF{$is(node, VarRefNode)$}
	\IF{$node.varName$ == $varName$}
		\STATE $type$ = $getType(node)$
		\IF {$type$ is $PathType$ AND $isTargetPath(type, varName)$}
			\STATE add $type$ to $paths$
		\ENDIF
	\ENDIF
\ENDIF
\FORALL{$child \in node.getChildren()$}
	\STATE $getTargetReturnPathsRecursive(child, varName, paths)$
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

The first XQuery construct utilized by this step is a function call of $distinct-values$ function as discussed in Section \ref{Negative_uniqueness} and shown in Algorithm \ref{ALG_rejection_of_uniqueness_aggregation_functions}. The algorithm searches for calls of the mentioned functions with a $PathType$ on their input. Elements and attributes selected by those $PathTypes$ are then considered to be not unique, and consequently, they cannot be keys.

Aggregation functions $min$, $max$, $sum$ are not utilized by the function because they were utilized by the search for join patterns, and in the process of classification of found join patterns, they were included in the computation of the key weights.

@@

\IF{$is(node, FunctionCallNode)$}
    \IF{$node.fncName$ == $distinct-values$}
        \STATE $argument = node.getChild(0)$
        \STATE $type = getType(argument)$
        \IF{$type$ is $PathType$ AND $usesOnlyChildAndDescendantAxes(type)$ AND $isWithoutPredicatesExceptLastStep(type)$}
            \STATE memorize $argument$ as the negative key statement, weight 1
        \ENDIF
    \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}
